---
title: ''
output:
  html_document:
    fig_height: 4
    fig_width: 5
    highlight: textmate
    theme: cerulean
  word_document: default
---

```{r setup, echo=FALSE, message=F, warning=F  }

#==============================================================================
#   This chunk is for every assignment: Modify line #24, then skip to line #79
#==============================================================================

  # Clear the working space
    rm(list = ls())
  
  # Change just this working directory command line (#23) and then skip  down
  setwd(yourworkingdirectoryhere))

### Load the packages (all must have been installed)
    
    library(AER)
    library(car)
    library(countrycode)
    library(doBy)
    library(dplyr)
    library(foreign)
    library(gdata)
    library(ggplot2)
    library(knitr)
    library(lmtest)
    library(readstata13)
    library(reshape)
    library(sandwich)
    library(stargazer)
    library(WDI)
    library(XML)

  ### More settings
    
  # Root directory for knitr
    opts_knit$set(root.dir = getwd())
  # turn off scientific notation except for big numbers
    options(scipen = 9)
  
  # function to calculate corrected SEs for OLS regression 
    cse <- function(reg) {
      rob = sqrt(diag(vcovHC(reg, type = "HC1")))
      return(rob)
    }
  
```

<base target="_top"/>
  
ECON 42  

Lab Assignment #1

Bill

 

*** 

## Getting started with R and R Markdown 
  
### 1. Introduction 
  
This document is a _script_ for __R Markdown__, which you are editing with __RStudio__. It contains instructions to create a document along with "chunks" of code that will execute commands in the statistical program __R__. When you run it by hitting the "Knit HTML" button, __R__ will load the data and crunch the numbers per your instructions, and R Markdown will incorporate the results with your write-up into a single report. _This means your work is 100% reproducible._ To do your labs, you will edit and run scripts like this one and submit both the script and the HTML file containing the results. 

Scroll down and take a look through the whole script. Don't worry if a lot of it seems mysterious right now. 

The script contains both text and lines of R code to run the statistical analysis. The lines of R code are in blocks or "chunks" which begin and end with three single back quotes:  \```\. In your editor the R chunks can be distinguished by a light gray background. The text can be formatted in a variety of ways: _italics_, __bold__, etc. You'll see how it works when you run it.

### 2. Examples 

Let's look at some chunks of R code that do the following:  

* Read in a data set on California school districts and create a table of descriptive statistics. We'll be using these data throughout the quarter.  
* Create a scatter plot of average test scores against the natural log of average annual per capita income, where each circle represents a school district.  
* Run two regressions and report the results in a table.  

At the beginning of the chunk in the script you will see the option ```echo=T```. That means that the code itself will be printed in your document, along with the results. At least for now, I want you to do this in your assignments, so I can see your code. Later in the course we can set ```ECHO=F```, which will make your homework reports much more readable. 

#### Descriptive statistics

Here is the code for a table of descriptive statistics followed by the table itself: 

```{r, message=F, warning=F, comment=NA, echo=T}

#==============================================================================
#   Data section
#==============================================================================

  # Data input using read.dta (from Stock & Watson textbook site)
  caschool <- read.dta("http://wps.aw.com/wps/media/objects/11422/11696965/datasets3e/datasets/caschool.dta")

  # Create a new variable
  caschool$loginc <- log(caschool$avginc)

#==============================================================================
#   Analysis section
#==============================================================================
  
  # table of descriptive statistics for quantitative variables
  stargazer(caschool, 
            title="Table 1: CA school data", type="text", digits=2 )
  
```

*** 

#### Plot 

Here is the code for a plot of the average test score against the log of average income in 420 California school districts, followed by the plot:

```{r, message=F, warning=F, comment=NA, echo=T}

  # scatter plot of test score against log of income in district
  ggplot(caschool, aes(x=loginc, y=testscr)) + 
    labs(y = "Test score", x = "Log average income" , title = "Figure 1: CA test scores and district income") +
    geom_point(shape=1) +           # use hollow circles
    geom_smooth(method=lm, se=F)    # add regression line

```

*** 

#### Regressions

Here is the code to run two regressions and display the results in a table, followed by the table. We'll be learning all about this! 

```{r, message=F, warning=F, comment=NA, echo=T}

  # regressions
  reg1 <- lm(testscr ~ avginc , data = caschool)  
  reg2 <- lm(testscr ~ avginc + str , data = caschool) 
  
  # regression table
  stargazer(reg1, reg2, 
            se=list(cse(reg1), cse(reg2)),
            title="Table 2: Test score regressions", type="text", 
            column.labels=c("Simple", "Multiple"),
            df=FALSE, digits=2)

``` 
